---
title: "Final Project"
author: "Hayden Beyer"
format: 
  html:
    embed-resources: true
---

## Abstract
- The Himalayan Database is a comprehensive archive documenting mountaineering expeditions in the Nepal Himalaya. This data is rich in historical value, detailing the peaks, expeditions, climbing statuses, and geographic information of numerous Himalayan summits. I will explore these data in two tidy tibbles, making it easier to analyze trends in mountaineering expeditions, including seasonality, success rates, and national participation over time. To manage  size, the expeditions file was filtered down to the years 2020-2024.

# Data Read
```{r message=FALSE}
library(tidyverse)
exped_tidy <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-01-21/exped_tidy.csv')
peaks_tidy <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-01-21/peaks_tidy.csv')
# peaks_tidy$PKNAME
# mean(peaks_tidy$HEIGHTM)
# mean(peaks_tidy$HEIGHTF)
```

# Top 10 Most Popualar Peaks
```{r}
exped_rank <- exped_tidy |> group_by(PEAKID) |> summarise(count = n()) |> ungroup() |> arrange(desc(count)) |> slice(1:10)

ggplot(exped_rank, aes(x = count, y = fct_reorder(PEAKID, count))) +
  geom_col() +
  labs(
    title = "Top 10 Most Popular Peaks",
    subtitle = "Based on total expedition counts",
    x = "Number of Expeditions",
    y = "Peak ID"
  ) +
  theme_minimal()
```

# Top 10 Most Popualar Peaks
```{r message=FALSE}
exped_nation <- exped_tidy |> group_by(NATION) |> summarise(count = n()) |> ungroup() |> arrange(desc(count)) |> slice(1:10)


ggplot(exped_nation, aes(x = count, y = fct_reorder(NATION, count))) +
  geom_col() +
  labs(
    title = "Top 10 Most Popular Nations",
    subtitle = "Based on total expedition counts",
    x = "Count",
    y = "Nations"
  ) +
  theme_minimal()
```

# Avg Range Height
```{r message=FALSE}
range_avgheight <- peaks_tidy |> group_by(HIMAL_FACTOR) |> summarise(mean_height = mean(HEIGHTF)) |> ungroup() |> arrange(desc(mean_height)) 

ggplot(range_avgheight, aes(x = mean_height, y = fct_reorder(HIMAL_FACTOR, mean_height))) +
  geom_col() +
  labs(
    title = "AVG Mountain Range Height (FT)",
    x = "AVG Height",
    y = "Ranges"
  ) +
  theme_minimal()
```


## NUM Deaths and NoHired
```{r message=FALSE}
exped_tidy |> group_by(HDEATHS, MDEATHS) |> summarise(no_deaths = n())
  
exped_tidy |> group_by(NOHIRED) |> summarise(count = n())

unique(exped_tidy$TERMREASON_FACTOR)
```


# Joining the datasets for models
```{r message=FALSE}
library(tidyverse)
model_df <- exped_tidy |>
  left_join(peaks_tidy, by = "PEAKID") |>
  mutate(
    success_binary = factor(ifelse(SUCCESS1, "Success", "Failure")),
    O2USED = factor(O2USED),
    SEASON_FACTOR = factor(SEASON_FACTOR)
  ) |>
  filter(!is.na(success_binary)) |> filter(!is.na(success_binary), !is.na(O2USED), !is.na(TOTHIRED)) |>
  mutate(hired_cat = case_when(
    TOTHIRED == 0  ~ "None (0)",
    TOTHIRED <= 5  ~ "Low (1-5)",
    TOTHIRED <= 15 ~ "Medium (6-15)",
    TOTHIRED > 15  ~ "High (15+)"
  ),
  # Ensure they appear in the correct order in the plot
  hired_cat = fct_relevel(hired_cat, "None (0)", "Low (1-5)", "Medium (6-15)", "High (15+)"))


```


# Number of hired personnel (above base camp) vs Success Rate
```{r message=FALSE}
library(tidyr)
library(broom)

success_glm <- glm(success_binary ~  HEIGHTM + O2USED +  TOTHIRED, family = "binomial", data = model_df)

success_glm |> tidy()

model_df <- model_df |>
  mutate(success_num = as.numeric(success_binary) - 1)

success_glm <- glm(success_num ~ HEIGHTM + O2USED + TOTHIRED,
                   family = "binomial", data = model_df)

library(ggeffects)
ggpredict(success_glm, terms = c("HEIGHTM", "O2USED", "TOTHIRED"))

preds_success <- ggpredict(success_glm, terms = c("HEIGHTM [all]", "O2USED", "TOTHIRED")) |>
  as.data.frame(terms_to_colnames = TRUE) 


ggplot(data = model_df, aes(x = HEIGHTM, y = as.numeric(success_binary) - 1)) +
  geom_jitter(height = 0.05) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  scale_y_continuous(breaks = c(0, 1), labels = c("Failure", "Success")) +
  labs(y = "success_binary") +
  theme_minimal()


ggplot(data = model_df, aes(x = TOTHIRED, y = as.numeric(success_binary) - 1)) +
  geom_jitter(height = 0.05) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  scale_y_continuous(breaks = c(0, 1), labels = c("Failure", "Success")) +
  labs(y = "success_binary") +
  theme_minimal()
```


# Probability of Success Base of Hired No. Personel, Peak Height and Oxygen Use. 
```{r message=FALSE}
library(tidyr)
library(broom)
library(ggeffects)
library(ggplot2)

# 1. Fit the model
success_glm <- glm(success_binary ~ HEIGHTM + O2USED + hired_cat, 
                   family = "binomial", 
                   data = model_df)

# 2. Generate predictions (Pipe ends here!)
preds_success <- ggpredict(success_glm, terms = c("HEIGHTM [all]", "O2USED", "hired_cat")) |>
  as.data.frame() # Removed the trailing |> 

# 3. Plot
# Note: x, y, and group names from ggpredict are often 'x', 'predicted', 'group', and 'facet'
# depending on your ggeffects version. Check colnames(preds_success) if this errors.

ggplot(data = preds_success, aes(x = x, y = predicted)) + 
  geom_line(aes(colour = group), linewidth = 1.5) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.3) + 
  # Use the original data for points, but force the factor to 0/1
  geom_jitter(data = model_df, 
              aes(x = HEIGHTM, y = as.numeric(success_binary) - 1, colour = O2USED), 
              width = 0, height = 0.04, alpha = 0.3) + 
  facet_wrap(~ facet) + 
  scale_colour_viridis_d() + 
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(x = "Peak Height (m)", y = "Probability of Success", colour = "Oxygen Used", fill = "Oxygen Used")

```


# GLM Probability of Success
```{r message=FALSE}
success_glm <- glm(success_binary ~ HEIGHTM + O2USED + hired_cat + SEASON_FACTOR + TOTMEMBERS, 
                   family = "binomial", 
                   data = model_df)

print(success_glm)
```









# Linear Model
```{r message=FALSE}
# Filter for successful expeditions only to model time-to-summit
duration_df <- model_df %>%
  filter(SUCCESS1 == TRUE, SMTDAYS > 0, SMTDAYS < 100)

# Fit the model
duration_fit <- lm(SMTDAYS ~ HEIGHTM + CAMPS + ROPE + SEASON_FACTOR, data = duration_df)

duration_fit |> tidy()


# Visualize the effect of Peak Height on Time to Summit
ggplot(duration_df, aes(x = HEIGHTM, y = SMTDAYS)) +
  geom_jitter(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", color = "darkred") +
  labs(title = "Impact of Peak Height on Days to Summit",
       x = "Peak Height (meters)",
       y = "Days to Reach Summit") +
  theme_minimal()
```



