---
title: "Himalayan Expeditions"
format: gfm
---

### Data
The Himalayan Database is a detailed archive documenting mountaineering expeditions in the Himalayans. This data details the peaks, expeditions, climbing statuses, and geographic information of numerous Himalayan summits. I will explore these data in two tidy tibbles, making it easier to analyze trends in mountaineering expeditions, including seasonality, success rates, and national participation over time. To manage size, the expeditions file was filtered down to the recent years of 2020-2024.

```{r message=FALSE}
library(tidyverse)
exped_tidy <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-01-21/exped_tidy.csv')
peaks_tidy <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-01-21/peaks_tidy.csv')
```


### Questions
- Is there a correlation between the number of deaths among hired staff versus expedition members, and which group typically has a higher mortality?

- How does the probability of expedition success change as peak height increases, and to what extent do supplemental oxygen and different levels of hired support help the as mountains go to higher elevations?


# Death Heatmaps
```{r}
deaths_df <- exped_tidy |>
  group_by(HDEATHS, MDEATHS) |>
  summarise(no_deaths = n(), .groups = "drop") |>
  mutate(
    HDEATHS = factor(HDEATHS),
    MDEATHS = factor(MDEATHS)
  )

ggplot(deaths_df, aes(x = HDEATHS, y = MDEATHS, fill = no_deaths)) +
  geom_tile(colour = "darkgray", linewidth = 0.8) +
  geom_text(aes(label = no_deaths), colour = "gray", fontface = "bold", size = 4) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    trans = "log10",        
  ) +
  labs(
    title = "Expedition Fatalities: Hired Staff vs. Members",
    x = "Hired Staff Deaths",
    y = "Member Deaths",
    fill = "# of Exped."
  ) +
  theme_minimal()
# AI was used in this graph to add log scale. Question: How do I scale this down?
```

The majority of expeditions (844 of 882) result in zero deaths of any kind. Deaths among hired staff and members tend to occur independently. Events with multiple fatalities involving both groups at the same time are extremely rare. Instead, they are usually isolated incidents involving one or two individuals, with expedition members appearing to die at a slightly higher rate in single death scenarios than the hired staff. This makes sense as the hired staff tends to be more prepared and experienced, however, accidents can still happen. 


# Probability of Success Base of Hired No. Personel, Peak Height and Oxygen Use. 
```{r message=FALSE}
library(tidyverse)
model_df <- exped_tidy |>
  left_join(peaks_tidy, by = "PEAKID") |>
  mutate(
    success_binary = factor(ifelse(SUCCESS1, "Success", "Failure")),
    O2USED = factor(O2USED),
    SEASON_FACTOR = factor(SEASON_FACTOR)
  ) |>
  filter(!is.na(success_binary)) |> filter(!is.na(success_binary), !is.na(O2USED), !is.na(TOTHIRED)) |>
  mutate(hired_cat = case_when(
    TOTHIRED == 0  ~ "None (0)",
    TOTHIRED <= 5  ~ "Low (1-5)",
    TOTHIRED <= 15 ~ "Medium (6-15)",
    TOTHIRED > 15  ~ "High (15+)"
  ),
  hired_cat = fct_relevel(hired_cat, "None (0)", "Low (1-5)", "Medium (6-15)", "High (15+)"))
```

```{r message=FALSE, results = 'hide'}
library(tidyr)
library(broom)
library(ggeffects)
library(ggplot2)


success_glm <- glm(success_binary ~ HEIGHTM + O2USED + hired_cat, 
                   family = "binomial", 
                   data = model_df)


preds_success <- ggpredict(success_glm, terms = c("HEIGHTM [all]", "O2USED", "hired_cat")) |>
  as.data.frame()


```

```{r message=FALSE, }
ggplot(data = preds_success, aes(x = x, y = predicted)) + 
  geom_line(aes(colour = group), linewidth = 1.5) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.3) + 
  geom_jitter(data = model_df, 
              aes(x = HEIGHTM, y = as.numeric(success_binary) - 1, colour = O2USED), 
              width = 0, height = 0.04, alpha = 0.3) + 
  facet_wrap(~ facet) + 
  scale_colour_viridis_d() + 
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(x = "Peak Height (m)", y = "Probability of Success", colour = "Oxygen Used", fill = "Oxygen Used")
```
In every scenario, the probability of success drops as Peak Height increases. This drop is most apparent for climbers not using supplemental oxygen (the dark purple lines). Comparing the four facets, None (0), Low (1-5), Medium (6-15), and High (15+), shows that increasing the number of hired personnel flattens the curve going down. It is interesting to point out that with High (15+) hired support, those not using oxygen still have a higher success rate at around 8,000 meters than those with no support at the same height. Generally speaking the gap between oxygen users and non users only really becomes a critical point as expeditions move into the death zone (above 7,500–8,000 meters).



